Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-20 introduction.tex
Christine Simpson's changes.

The \enzo\ code has been extensively used over the last two decades in
a wide variety of problems, resulting in the publication of more than
100 peer-reviewed papers based on results obtained with the code. The
variety of astrophysical systems that \enzo\ has been used for include
galaxies \citep{2003ApJ...587...13T, 2012MNRAS.425..641L, 2012ApJ...749..140H}, galaxy
clusters \citep{Loken02, Xu11, Skillman13}, the interstellar medium
\citep{Slyz05}, the intergalactic medium \citep{Fang01,
2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S},
turbulence \citep{Kritsuk04, 2007ApJ...665..416K, 2009JPhCS.180a2020K, Collins11}, the
formation of the first stars \citep{ABN02, 2007ApJ...654...66O,
2009Sci...325..601T, Xu08}, and the formation of stars in our own Galaxy
\citep{Collins11, 2011ApJ...727L..20K, Collins12a}.

>>>>

The \enzo\ code has been extensively used over the last two decades in
a wide variety of problems, resulting in the publication of more than
100 peer-reviewed papers based on results obtained with the code. The
variety of astrophysical systems that \enzo\ has been used for include
galaxies \citep{2003ApJ...587...13T, 2012MNRAS.425..641L,
  2012ApJ...749..140H, 2013MNRAS.432.1989S}, galaxy
clusters \citep{Loken02, Xu11, Skillman13}, the interstellar medium
\citep{Slyz05}, the intergalactic medium \citep{Fang01,
2011ApJ...731....6S}, cooling flows \citep{Li12, 2013ApJ...763...38S},
turbulence \citep{Kritsuk04, 2007ApJ...665..416K, 2009JPhCS.180a2020K, Collins11}, the
formation of the first stars \citep{ABN02, 2007ApJ...654...66O,
2009Sci...325..601T, Xu08}, and the formation of stars in our own Galaxy
\citep{Collins11, 2011ApJ...727L..20K, Collins12a}.

### 2013-06-20 numerical-amr.tex
Jeff Oishi's changes.
Christine Simpson's changes.

From the point of view of the hydro solver, each grid is solved as an independent CFD problem, with Dirichlet boundary conditions stored in the ghost zones.  At the end of each timestep, cells on a given level fill ghost zones by interpolating from the parent and copying from neighbors on the same level.  Grids then correct the flux at the interface boundary between fine and coarse zones, and finally project their active zone data to the parent grid.

@@

\item{\em Resistive length.}  In a way similar to the Jeans length criterion described above, the code can be forced to locally resolve the resistive length scale by a minimum of a fixed number of cells.  This is done by ensuring that the resistive length scale, defined as
\begin{equation}
L_R = |B| / |\curl B|
\end{equation}
is always refined by at least N cells locally (i.e.~we refine a given cell further if $L_R \leq N \Delta x$).

>>>>

From the point of view of the hydro solver, each grid is solved as an independent computational fluid dynamics (CFD) problem, with Dirichlet boundary conditions stored in the ghost zones.  At the end of each timestep, cells on a given level fill ghost zones by interpolating from the parent and copying from neighbors on the same level.  Grids then correct the flux at the interface boundary between fine and coarse zones, and finally project their active zone data to the parent grid.

@@

\item{\em Resistive length.}  In a way similar to the Jeans length criterion described above, the code can be forced to locally resolve the resistive length scale by a minimum of a fixed number of cells.  This is done by ensuring that the resistive length scale, defined as
\begin{equation}
L_R = |B| / |\curl B|
\end{equation}
is always refined by at least N cells locally (i.e.~we refine a given cell further if $L_R \leq N \Delta x$).  Note that \enzo\ does not actually solve the resistive MHD equations -- this is the ``characteristic'' local resistive length, rather than one calculated within the MHD solver.

### 2013-06-20 overview.tex
Remove \rho from EOS definition and noted that units are such that magnetic permeability is unity.
Jeff Oishi's changes.
Christine Simpson's changes.

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \grad \rho
%    & = & - \rho \div \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \grad ) \vecv 
%    & = & - \frac{1}{\rho} \grad p
%          - \grad \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \grad E
%    & = & - \frac{1}{\rho} \div (p \vecv) 
%          - \vecv \cdot \grad \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
In these equations, $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving
total fluid energy density, comoving gas density, peculiar
velocity, and comoving magnetic field strength, respectively.
$\myvec{I}$ is the identity matrix, and $a$ is the cosmological
expansion parameter (discussed in more detail below).  The first
equation represents conservation of mass, the second conservation of
momentum, and the third conservation of total (kinetic plus thermal
plus magnetic)
fluid energy.  They are respectively, the first, second, and third
moments of the Boltzmann equation.  The fourth equation is the
magnetic induction equation.  Terms representing radiative cooling
($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the
energy equation (\ref{eq:total_energy}), as does the flux due to
thermal heat conduction ($\fcond$).

The comoving total fluid energy density $E$ is given by
\begin{equation}
E =  e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def},
\end{equation}
where $e$ is the comoving thermal energy density, the total comoving
isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a},
\end{equation}
and $p$ is the thermal pressure.  The equations are closed by an
equation of state and Poisson's equation (in the Newtonian limit) for
the gravitational potential $\phi$:
%
\begin{eqnarray}
%
  e &=& \frac{p}{(\gamma - 1) \rho},
  \label{eq:eq_of_state} \\
%
  \nabla^2 \phi &=& 4 \pi G \left( \rho_{\rm total} - \rho_0 \right)
%                   + 3 p_{total}/c^2) - \Lambda. 
  \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of
specific heats $\gamma$.  The gravitational potential $\phi$ is
sourced by the total mass density contrast, where $\rho_{\rm total} =
\rho_{\rm gas} + \rho_{\rm dm} + \rho_{\rm stars}$ and $\rho_0$ is the
mean density.  Although we write the equations including both magnetic
field terms and comoving coordinates, the code is frequently used both
in the purely hydrodynamic limit ($\vecB = 0$, referred to as HD
below) and without cosmological expansion ($a = 1$, $\dot{a} = 0$).

@@

As an alternative to the ray-casting strategy, \enzo\ can also solve
the angle-averaged radiative transfer equation, through a
flux-limited diffusion approximation with couplings to both the gas
energy and chemical number densities. The equations solved are
\begin{eqnarray}
  \label{eq:fld_radiation}
  \frac{\partial E_r}{\partial t} + \frac1a \div\left(E_r \vecv \right) &=& 
  \div\left(D\grad E_r\right) -
  \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c + \Gamma - \Lambda,
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \vecv_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r = E_r(\vecx,t)$ is a grey radiation energy density and
$e_c$ is the internal energy correction due to photo-heating and
chemical cooling.
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  
Here, we define $E_r$ through first assuming a fixed frequency
spectrum, i.e.  $E_{\nu}(\nu,\vecx,t) = \tilde{E}_r(\vecx,t)
\chi(\nu)$, and then defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vecx,t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vecx,t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vecx,t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
$D$ in Equation (\ref{eq:fld_radiation}) is the Larsen
square-root flux-limiter \citep[see][]{Morel2000}, $\kappa$ the
total opacity, $\eta$ the field of radiation sources, $\Gamma$
the radiation induced photo-heating rate, and $\Lambda$ the
chemical cooling rate.
%, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

@@

The primary purpose of the \enzo\ code is its Adaptive Mesh Refinement
(AMR) capability, which allows it to reach extremely large dynamical
ranges with limited computational resources, opening doors otherwise
closed by finite memory and computational time. Unlike moving mesh
methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or methods
that subdivide individual cells \citep{Adjerid}, Berger \& Collela's
AMR (also referred to as \emph{structured} AMR) utilizes an adaptive
hierarchy of grid patches at varying levels of resolution.  Each
rectangular grid patch (referred to as a ``grid'') covers some region
of space in its \emph{parent grid} requiring higher resolution,
and can itself become the parent grid to an even more highly resolved
\emph{child grid}.

The grid hierarchy begins with the root grid, which covers the entire
domain of interest with a coarse uniform Cartesian grid. Then, as
the solution evolves and interesting regions start to develop, finer meshes are
placed within these coarse regions.
% (we use the notation `below' to refer to
% finer grids and `above' for coarser grids). 
We restrict the ratio between cell sizes to be an integer, typically 2
or 4, and refer to a level as all grids with the same cell size.  In
order to keep things as simple as possible, the edges of subgrids must
coincide with the cell edge of its immediate parent (coarser)
grid. Additionally, the hierarchy can be initialized with one or more
static grids if a higher initial resolution is required.

@@

Since interesting regions on the grid may move, the hierarchy must
adapt itself. This happens whenever a level has caught up to its
parent level, by entirely rebuilding the grids on that level and at
finer resolutions. Rebuilding is achieved by applying the grid
refinement criteria to the grids on that level, flagging zones that
require extra grids.  These criteria depend on the physical problem
being simulated; see Section~\ref{sec:refinement_criteria} for more
details.  Once a grid has a set of flagged cells, we apply a
machine-vision based algorithm \citep{Berger91} to find edges and
determine a good placement of subgrids.
%  These subgrids must not overlap one another, must cover all flagged
% cells and their neighbouring cells, and be above a preset efficiency
% threshold, where the efficiency of a cell is defined as the ratio of
% flagged cells to total cells. 
Once these new subgrids have been identified, the solution from the
next coarser grid is interpolated in order to initialize the values on
the new grids.  Finally, any overlap between new subgrids and old ones
is identified, and the prior solution within the regions of overlap is
copied to the new subgrids. This procedure is then repeated on the new
grids and in this way the entire hierarchy (from the original level
examined and continuing on to all finer levels) is rebuilt.

@@

\enzo\ can operate in a number of different modes with regard to
radiative cooling and heating. In the simplest mode, with the
multi-species flag turned off, the cooling rate is computed from a
simple temperature-dependent cooling function, taken from
\citet{SW87}.  If chemistry is turned on, then the code can include
cooling from all species of hydrogen and helium (including H$_2$) --
this primordial cooling is treated in the same Jacobi iteration as the
chemistry.  It is also possible to include metal cooling based on a
set of multi-dimensional lookup tables computed with the
\textsc{Cloudy} code \citep{1998PASP..110..761F} as described in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}. Note that
the cooling and heating is most commonly treated in the optically-thin
limit, but the code can also follow radiative transfer in a limited
set of energy bins.

@@

All grids on a given level are advanced with the same timestep,
determined by calculating, for each cell and for each physical process
(except for chemistry and heat conduction, which are sub-cycled)
separately, the largest $\Delta t$ allowed for stability. The level is
advanced with a timestep equal to the minimum over all of these
$\Delta t$.

>>>>

%\begin{eqnarray} 
%\frac{\partial \rho}{\partial t} 
%          + \vecv \cdot \grad \rho
%    & = & - \rho \div \vecv,
%        \label{eq:mass} \\
%%
%\frac{\partial \vecv }{ \partial t} 
%          + ( \vecv \cdot \grad ) \vecv 
%    & = & - \frac{1}{\rho} \grad p
%          - \grad \phi,
%        \label{eq:momentum} \\
%%
% \frac{\partial E}{\partial t} + \vecv \cdot \grad E
%    & = & - \frac{1}{\rho} \div (p \vecv) 
%          - \vecv \cdot \grad \phi
%%\nonumber          \\  & &     % uncomment for 2-column
%     + \Gamma - \Lambda
%        \label{eq:total_energy}
%\end{eqnarray}
%
In these equations, $E$, $\rho$, $\vecv$, and $\vecB$ are the comoving
total fluid energy density, comoving gas density, peculiar
velocity, and comoving magnetic field strength, respectively. The
matrix 
$\myvec{I}$ is the identity matrix, and $a$ is the cosmological
expansion parameter (discussed in more detail below).  The first
equation represents conservation of mass, the second conservation of
momentum, and the third conservation of total (kinetic plus thermal
plus magnetic) fluid energy.  They are respectively, the first, second, and third
moments of the Boltzmann equation.  The fourth equation is the
magnetic induction equation.  Terms representing radiative cooling
($\Lambda$) and heating ($\Gamma$) enter on the right-hand side of the
energy equation (\ref{eq:total_energy}), as does the flux due to
thermal heat conduction ($\fcond$).

The comoving total fluid energy density $E$ is given by
\begin{equation}
E =  e + \frac{\rho v^2}{2}  + \frac{B^2}{2a}
        \label{eq:total_energy_def},
\end{equation}
where $e$ is the comoving thermal energy density. The total comoving
isotropic pressure $p^*$ is given by
\begin{equation}
p^* = p + \frac{B^2}{2a},
\end{equation}
and the quantity $p$ is the thermal pressure.  We use units such that the magnetic
permeability is unity ($\mu_0=1$).  The equations are closed by an
equation of state and Poisson's equation (in the Newtonian limit) for
the gravitational potential $\phi$:
%
\begin{eqnarray}
%
  e &=& \frac{p}{(\gamma - 1)},
  \label{eq:eq_of_state} \\
%
  \nabla^2 \phi &=& 4 \pi G \left( \rho_{\rm total} - \rho_0 \right)
%                   + 3 p_{total}/c^2) - \Lambda. 
  \label{eq:potential}
\end{eqnarray}
%
The equation of state is shown here for an ideal gas with a ratio of
specific heats $\gamma$.  The gravitational potential $\phi$ is
sourced by the total mass density contrast, where $\rho_{\rm total} =
\rho_{\rm gas} + \rho_{\rm dm} + \rho_{\rm stars}$ and $\rho_0$ is the
mean density.  Although we write the equations including both magnetic
field terms and comoving coordinates, the code is frequently used both
in the purely hydrodynamic limit ($\vecB = 0$, referred to as HD
below) and without cosmological expansion ($a = 1$, $\dot{a} = 0$).

@@

As an alternative to the ray-casting strategy, \enzo\ can also solve
the angle-averaged radiative transfer equation, through a
flux-limited diffusion approximation with couplings to both the gas
energy and chemical number densities. The equations solved are
\begin{eqnarray}
  \label{eq:fld_radiation}
  \frac{\partial E_r}{\partial t} + \frac1a \div\left(E_r \vecv \right) &=& 
  \div\left(D\grad E_r\right) -
  \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c + \Gamma - \Lambda,
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \div\left({\mathrm n}_i \vecv_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r = E_r(\vecx,t)$ is a grey radiation energy density and
$e_c$ is the internal energy correction due to photo-heating and
chemical cooling.
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  
Here, we define $E_r$ through first assuming a fixed frequency
spectrum, i.e.  $E_{\nu}(\nu,\vecx,t) = \tilde{E}_r(\vecx,t)
\chi(\nu)$, and then defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vecx,t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vecx,t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vecx,t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
The quantity $D$ in Equation (\ref{eq:fld_radiation}) is the Larsen
square-root flux-limiter \citep[see][]{Morel2000}, $\kappa$ is  the
total opacity, $\eta$ is the field of radiation sources, $\Gamma$ is
the radiation induced photo-heating rate, and $\Lambda$ is the
chemical cooling rate.
%, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

@@

The primary feature of the \enzo\ code is its Adaptive Mesh Refinement
(AMR) capability, which allows it to reach extremely large dynamical
ranges with limited computational resources, opening doors to applications otherwise
closed by finite memory and computational time. Unlike moving mesh
methods \citep{1995ApJS..100..269P,1995ApJS...97..231G} or methods
that subdivide individual cells \citep{Adjerid}, Berger \& Collela's
AMR (also referred to as \emph{structured} AMR, or SAMR) utilizes an adaptive
hierarchy of grid patches at varying levels of resolution.  Each
rectangular grid patch (referred to as a ``grid'') covers some region
of space in its \emph{parent grid} requiring higher resolution,
and can itself become the parent grid to an even more highly resolved
\emph{child grid}.

The grid hierarchy begins with the root grid, which covers the entire
domain of interest with a coarse uniform Cartesian grid. Then, as
the solution evolves and interesting regions start to develop, finer grids are
placed within these coarse regions.
% (we use the notation `below' to refer to
% finer grids and `above' for coarser grids). 
We restrict the ratio between cell sizes of parent and child grids to be an integer, typically 2
or 4, and refer to a level as all grids with the same cell size.  For simplicity, the edges of subgrids must
coincide with the cell edge of its immediate parent (coarser)
grid. Additionally, the hierarchy can be initialized with one or more
static grids if a higher initial resolution is required.

@@

Since interesting regions on the grid may move, the hierarchy must
adapt itself. This happens whenever a level has caught up to its
parent level, by entirely rebuilding the grids on that level and at
finer resolutions. Rebuilding is achieved by applying the grid
refinement criteria to the grids on that level and flagging zones that
require extra grids.  These criteria depend on the physical problem
being simulated; see Section~\ref{sec:refinement_criteria} for more
details.  Once a grid has a set of flagged cells, we apply a
machine-vision based algorithm \citep{Berger91} to find edges and
determine a good placement of subgrids.
%  These subgrids must not overlap one another, must cover all flagged
% cells and their neighbouring cells, and be above a preset efficiency
% threshold, where the efficiency of a cell is defined as the ratio of
% flagged cells to total cells. 
Once these new subgrids have been identified, the solution from the
next coarser grid is interpolated in order to initialize the values on
the new grids.  Finally, any overlap between new subgrids and old ones
is identified, and the prior solution within the regions of overlap is
copied to the new subgrids. This procedure is then repeated on the new
grids and in this way, the entire hierarchy (from the original level
examined and continuing on to all finer levels) is rebuilt.

@@

\enzo\ can operate in a number of different modes with regard to
radiative cooling and heating. In the simplest mode, where the
multi-species flag is turned off and no individual chemical species
are tracked, the cooling rate is computed from a
simple temperature-dependent cooling function, taken from
\citet{SW87}.  If chemistry is turned on, then the code can include
cooling from all species of hydrogen and helium (including H$_2$) --
and the primordial cooling rates are computed in the same Jacobi iteration as the
chemistry.  It is also possible to include metal cooling based on a
set of multi-dimensional lookup tables computed with the
\textsc{Cloudy} code \citep{1998PASP..110..761F} as described in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}. Note that
the cooling and heating is most commonly treated in the optically-thin
limit, but the code can also follow radiative transfer in a limited
set of energy bins.

@@

All grids on a given level are advanced with the same timestep.  This
time step is determined by first calculating the largest time step
allowed for each cell and for each physical process separately (except
for chemistry and heat conduction, which are sub-cycled).  The level
is then advanced with a timestep equal to the minimum over all of
these $\Delta t$.
